### 2021-01-08 ch01_introduction/introduction.tex
Cleaned code
Added figures for ESMValTool contributions (chapter 3)
Cleaned tex files and started chapter 04

- natural vs. anthropogenic effects (D\&A) -> clear evidence that climate
change over 21st century is caused by human influence
(https://globalwarmingindex.org/, Haustein et al., 2017)

@@

@@

\begin{enumerate}
  \item Did climate sensitivity change in the latest generation of climate
  models from \acs{CMIP}6 and can we see progress compared to older climate
  model generations?

@@

This thesis is structured as follows: \Cref{ch:02:scientific_background}
introduces the scientific background. This includes relevant literature that is
used as a baseline for this thesis. \Cref{ch:03:esmvaltool} gives an overview
over the contributions made to the \ac{ESMValTool}, an open-source software for
the analysis of \acp{ESM}. These contributions helped improving the routine
evaluation of \acp{ESM} which is useful for the entire scientific community and
lead to co-authorship in four peer-reviewed studies \autocite{Eyring2020,
  Lauer2020, Righi2020, Weigel2020}. \Cref{ch:04:assessment_climate_metrics}
covers the assessment of policy-relevant climate metrics like the \ac{ECS} and
the \ac{TCR} in the latest generation of \acp{ESM}. This work is already
published in two scientific publications \autocite{Bock2020, Meehl2020}. Since
the \ac{ECS} and \ac{TCR} are considerably higher in this new climate model
generation, \cref{ch:05:paper_ecs} describes the assessment of emergent
constraints (a technique to reduce uncertainties in multi-model climate
projections, see \cref{subsec:02:emergent_constraints}) on the \ac{ECS} for
these \acp{ESM}. The contents of this chapter are published in \emph{Earth
  System Dynamics} \autocite{Schlund2020a}. \Cref{ch:06:paper_gpp} focuses on a
new method to reduce uncertainties in gridded multi-model projections of the
future climate with observations based on \ac{ML}. As an example, the method is
applied to the photosynthesis rate at the end of the \nth{21} century, which is
already published in the \emph{Journal of Geophysical Research: Biogeosciences}
\autocite{Schlund2020}. Finally, \cref{ch:07:summary_outlook} provides a
summary of the results of this thesis and gives an outlook of possible future
works.

>>>>

- natural vs. anthropogenic effects (D\&A) -> clear evidence that climate
change over 21st century is caused by human influence. The following website:
is relevant: (https://globalwarmingindex.org/, Haustein et al., 2017)

@@

- Mention IPCC assessments, older ones and upcoming AR6

- "climate model" is used as generic term which include \acp{AOGCM} and
\acp{ESM} (see \cref{subsec:02:numerical_climate_modeling}) (subcategories).
Since most modern climate models participating in CMIP are \acp{ESM} (or at
least have \ac{ESM} versions), the terms "climate model" and "ESM" are used
interchangeably in this thesis.

@@

\begin{enumerate}
  \item Did climate sensitivity change in the latest generation of \acp{ESM} from \acs{CMIP}6 and can we see progress compared to older climate
  model generations?

@@

This thesis is structured as follows: \Cref{ch:02:scientific_background}
introduces the scientific background. This includes relevant literature that is
used as a baseline for this thesis. \Cref{ch:03:esmvaltool} gives an overview
over the contributions made to the \ac{ESMValTool}, an open-source software for
the analysis of \acp{ESM}. These contributions helped improving the routine
evaluation of \acp{ESM} which is useful for the entire scientific community and
lead to co-authorship in four peer-reviewed studies \autocite{Eyring2020,
  Lauer2020, Righi2020, Weigel2020}. \Cref{ch:04:assessment_climate_metrics}
covers the assessment of climate sensitivity metrics like the \ac{ECS} or
\ac{TCR} in the latest generation of \acp{ESM} (\acs{CMIP}6). This work is
already published in two scientific publications \autocite{Bock2020,
  Meehl2020}. Since the \ac{ECS} and \ac{TCR} are considerably higher in this
new climate model generation, \cref{ch:05:paper_ecs} describes the assessment
of emergent constraints (a technique to reduce uncertainties in multi-model
climate projections, see \cref{subsec:02:emergent_constraints}) on the \ac{ECS}
for these \acp{ESM}. The contents of this chapter are published in \emph{Earth
  System Dynamics} \autocite{Schlund2020a}. \Cref{ch:06:paper_gpp} focuses on a
new method to reduce uncertainties in gridded multi-model projections of the
future climate with observations based on \ac{ML}. As an example, the method is
applied to the photosynthesis rate at the end of the \nth{21} century, which is
already published in the \emph{Journal of Geophysical Research: Biogeosciences}
\autocite{Schlund2020}. Finally, \cref{ch:07:summary_outlook} provides a
summary of the results of this thesis and gives an outlook of possible future
works.

### 2021-01-08 ch03_esmvaltool/esmvaltool.tex
Cleaned tex files and started chapter 04

As shown in \cref{subsec:02:climate_modeling}, climate models have been
continuously improved and extended over the last decades from the relatively
simple \acp{AOGCM} to the complex state-of-the-art \acp{ESM} that include an
immense number of variables and processes. In \acs{CMIP}6, more modeling
institutes provide data for more versions of these complex models running
simulations for more experiments. Consequently, the data volume of the entire
\acs{CMIP}6 archive is expected to reach up to $80 \unit{PB}$
\autocite{Balaji2018}, which is a vast increase in comparison to the $2
\unit{PB}$ of \acs{CMIP}5. The increasing complexity and data volume of the
climate models pose a major challenge for the evaluation and analysis of the
model output. To address this big data challenge and support the scientific
community, the \ac{ESMValTool} has been developed to provide an
\enquote{open-source, standardized, community-based software package for the
  systematic, efficient, and well-documented analysis of \ac{ESM} results}
\autocite{Righi2020}. Moreover, the \ac{ESMValTool} allows a routine comparison
of single or multiple climate models against predecessor versions and/or
observations.

@@

Apart from new preprocessor functions, six derivation scripts for the
non-\ac{CMOR} variables \emph{asr}, \emph{co2s}, \emph{et}, \emph{rlntcs},
\emph{rsntcs} and \emph{uajet} have been added. The absorbed shortwave
radiation (\emph{asr}) is defined as the difference in the incoming \ac{TOA}
shortwave radiation and the outgoing \ac{TOA} shortwave radiation. The
atmospheric \ac{CO2} concentration at surface (\emph{co2s}) can be calculated
from the pressure level-dependent atmospheric \ac{CO2} concentration and the
surface air pressure using interpolation. An example illustrating \emph{co2s}
is given in \cref{fig:03:co2s}. \Cref{fig:03:co2s:a} shows the \acs{CMIP}6
\ac{MMM} of the global \emph{co2s} for the months \acf{JJA} of the year 2014 in
the emission-driven historical simulation. As expected, the map shows high
atmospheric \ac{CO2} concentrations over large metropolitan areas with high
\ac{CO2} emissions (\eg{} over North America, the Arabian Peninsula and East
Asia). Moreover, regions with high photosynthetic activity (\eg{} the boreal
forests in the Northern hemisphere summer and the tropical rainforests in South
America) exhibit smaller \ac{CO2} concentrations. \Cref{fig:03:co2s:b} shows
the monthly-mean \emph{co2s} at Mauna Loa, Hawaii from 1958 to 2014 for
observations performed at the Mauna Loa Observatory \autocite{Keeling2005}
(thick black line) and emission-driven historical simulations from four
\acs{CMIP}6 models (remaining colored lines). Apart from the model CNRM-ESM2-1,
the simulated atmospheric \ac{CO2} concentrations show the expected seasonal
cycle present in the Keeling Curve (see \cref{fig:02:keeling_curve}). All
climate models correctly simulate the increase in \emph{co2s} over the years.
However, there are some differences in the absolute values with the model
BCC-CSM2-MR showing the largest deviations.

@@

Due to an exhaustive overhaul of the \ac{ESMValTool} version 2, older recipes
from version 1 \autocite{Eyring2016a} cannot be used directly in the latest
release. Porting recipes from the old version to the current version requires
substantial changes on the recipes themselves, but also on the corresponding
diagnostic scripts. As of December 2020, the author of this thesis has ported
two recipes originally written by other authors to the new version of
\ac{ESMValTool}: \emph{recipe\_anav13jclim.yml} and
\emph{recipe\_wenzel16jclim.yml}. \emph{recipe\_anav13jclim.yml} reproduces the
analysis of \textcite{Anav2013}, who evaluate carbon cycle-related variables
for climate models of the \acs{CMIP}5 ensemble. This includes climatologies,
trends and variabilities of important quantities of the terrestrial and oceanic
carbon cycle like carbon fluxes, carbon reservoir sizes and vegetation
distributions. \emph{recipe\_wenzel16jclim.yml} includes the \ac{MDER} analysis
performed by \textcite{Wenzel2016a} to constrain future changes in the position
of the austral jet stream. Due to its flexible code, the corresponding
diagnostic scripts can be used to apply the \ac{MDER} method to arbitrary
target variables and predictors.

>>>>

As outlined in \cref{subsec:02:numerical_climate_modeling}, climate models have
been continuously improved and extended over the last decades from the
relatively simple \acp{AOGCM} to the complex state-of-the-art \acp{ESM} that
include an immense number of variables and processes. In \acs{CMIP}6, more
modeling institutes provide data for more versions of these complex models
running simulations for more experiments. Consequently, the data volume of the
entire \acs{CMIP}6 archive is expected to reach up to $80 \unit{PB}$
\autocite{Balaji2018}, which is a vast increase in comparison to the $2
\unit{PB}$ of \acs{CMIP}5. The increasing complexity and data volume of the
climate models pose a major challenge for the evaluation and analysis of the
model output. To address this big data challenge and support the scientific
community, the \ac{ESMValTool} has been developed to provide an
\enquote{open-source, standardized, community-based software package for the
  systematic, efficient, and well-documented analysis of \ac{ESM} results}
\autocite{Righi2020}. Moreover, the \ac{ESMValTool} allows a routine comparison
of single or multiple climate models against predecessor versions and/or
observations.

@@

Apart from new preprocessor functions, six derivation scripts for the
non-\ac{CMOR} variables \emph{asr}, \emph{co2s}, \emph{et}, \emph{rlntcs},
\emph{rsntcs} and \emph{uajet} have been added. The absorbed shortwave
radiation (\emph{asr}) is defined as the difference in the incoming \ac{TOA}
shortwave radiation and the outgoing \ac{TOA} shortwave radiation. The
atmospheric \ac{CO2} concentration at surface (\emph{co2s}) can be calculated
from the pressure level-dependent atmospheric \ac{CO2} concentration and the
surface air pressure using interpolation. An example illustrating \emph{co2s}
is given in \cref{fig:03:co2s}. \Cref{fig:03:co2s:a} shows the \acs{CMIP}6
\ac{MMM} of the global \emph{co2s} for the months \acf{JJA} of the year 2014 in
the emission-driven historical simulation. As expected, the map shows high
atmospheric \ac{CO2} concentrations over large metropolitan areas with high
\ac{CO2} emissions (\eg{} over North America, the Arabian Peninsula and East
Asia). Moreover, regions with high photosynthetic activity (\eg{} the boreal
forests in the Northern hemisphere summer and the tropical rainforests in South
America) exhibit smaller \ac{CO2} concentrations. \Cref{fig:03:co2s:b} shows
the monthly-mean \emph{co2s} at Mauna Loa, Hawaii from 1958 to 2014 for
observations performed at the Mauna Loa Observatory \autocite{Keeling2005}
(thick black line) and emission-driven historical simulations from four
\acs{CMIP}6 models (remaining colored lines). Apart from the model CNRM-ESM2-1,
the simulated atmospheric \ac{CO2} concentrations show the expected seasonal
cycle present in the Keeling Curve (see \cref{fig:02:keeling_curve}). All
\acp{ESM} correctly simulate the increase in \emph{co2s} over the years.
However, there are some differences in the absolute values with the model
BCC-CSM2-MR showing the largest deviations.

@@

Due to an exhaustive overhaul of the \ac{ESMValTool} version 2, older recipes
from version 1 \autocite{Eyring2016a} cannot be used directly in the latest
release. Porting recipes from the old version to the current version requires
substantial changes on the recipes themselves, but also on the corresponding
diagnostic scripts. As of December 2020, the author of this thesis has ported
two recipes originally written by other authors to the new version of
\ac{ESMValTool}: \emph{recipe\_anav13jclim.yml} and
\emph{recipe\_wenzel16jclim.yml}. \emph{recipe\_anav13jclim.yml} reproduces the
analysis of \textcite{Anav2013}, who evaluate carbon cycle-related variables
for \acp{ESM} of the \acs{CMIP}5 ensemble. This includes climatologies, trends
and variabilities of important quantities of the terrestrial and oceanic carbon
cycle like carbon fluxes, carbon reservoir sizes and vegetation distributions.
\emph{recipe\_wenzel16jclim.yml} includes the \ac{MDER} analysis performed by
\textcite{Wenzel2016a} to constrain future changes in the position of the
austral jet stream. Due to its flexible code, the corresponding diagnostic
scripts can be used to apply the \ac{MDER} method to arbitrary target variables
and predictors.

### 2021-01-08 ch04_assessment_climate_metrics/assessment_climate_metrics.tex
Initialized chapters
Wrote first page of introduction
Added first figure
Added section 2.2
Added overview of climate feedbacks
Added figures for ESMValTool contributions (chapter 3)
Cleaned tex files and started chapter 04

%=============================================================================%
%                                Dissertation                                 %
%                               Manuel Schlund                                %
%                                  (c) 2020                                   %
%=============================================================================%
%                                 ESMValTool                                  %
%=============================================================================%

\chapter{Improving routine Model Evaluation}
\label{ch:esmvaltool}

TBA.

>>>>

%=============================================================================%
%                                Dissertation                                 %
%                               Manuel Schlund                                %
%                                  (c) 2020                                   %
%=============================================================================%
%                        Assessment of climate metrics                        %
%=============================================================================%

\chapter{Assessment of Climate Sensitivity in \acs{CMIP}6}
\label{ch:04:assessment_climate_metrics}

In order to reduce uncertainties in multi-model climate projections, a first
important step is the assessment of the desired target variables in the
corresponding climate model ensemble. This is particularly relevant when a new
generation of \acp{ESM} has been published, which has happened in 2020 with the
\acs{CMIP}6 ensemble.
